Do researchers preferentially collaborate with colleagues of the same gender?

### 2018-06-03 Gender_homophily_analysis.Rmd
Plos references

### Gender homophily covaries with author gender ratio
Journals with a male-biased author gender ratio showed less author gender homophily, while journals with an even or female-biased author gender ratio showed more homophily. The pattern was similar whether we calculated $\alpha$ for all authors, first authors only, or last authors only. 
```{r fig.width=6, fig.height=5}
gender.ratio <- master.dataset %>% 
  filter(Age == "New") %>% 
  ggplot(aes(x = (gender.ratio.of.sample - 50), y = alpha.adjusted)) + 
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point(alpha = 0.4, size = 0.7) + 
  stat_smooth(method = "gam", 
              formula = y ~ s(x, k = 3), 
              colour = "tomato", fill = "skyblue") + 
  xlab("% women authors (realtive to gender parity)") + 
  ylab("Coefficient of homophily (\u03B1')") +
  facet_wrap(~Subset) + 
  theme_minimal() +
  theme(legend.position = "none") 
ggsave(plot = gender.ratio, filename = "figures/figure3.pdf", 
       device = cairo_pdf, width=7, height=5)
gender.ratio
```
<br></br>
**Figure `r figure.counter`**: There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3.

>>>>

### Gender homophily covaries with author gender ratio
Journals with a male-biased author gender ratio showed less author gender homophily, while journals with an even or female-biased author gender ratio showed more homophily. The pattern was similar whether we calculated $\alpha$ for all authors, first authors only, or last authors only. 
```{r fig.width=6, fig.height=5}
gender.ratio <- master.dataset %>% 
  filter(Age == "New") %>% 
  ggplot(aes(x = (gender.ratio.of.sample - 50), y = alpha.adjusted)) + 
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point(alpha = 0.4, size = 0.7) + 
  stat_smooth(method = "gam", 
              formula = y ~ s(x, k = 3), 
              colour = "tomato", fill = "skyblue") + 
  xlab("% women authors (relative to gender parity)") + 
  ylab("Coefficient of homophily (\u03B1')") +
  facet_wrap(~Subset) + 
  theme_minimal() +
  theme(legend.position = "none") 
ggsave(plot = gender.ratio, filename = "figures/figure3.pdf", 
       device = cairo_pdf, width=7, height=5)
gender.ratio
```
<br></br>
**Figure `r figure.counter`**: There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3.

### 2018-06-03 manuscript/Manuscript_Claire.Rmd
Plos references

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious
  biases, and often feel excluded from formal and informal opportunities for research
  collaboration. In addition to ensuring fairness and helping to redress gender imbalance
  in the academic workforce, increasing women's access to collaboration could help
  scientific progress by drawing on more of the available human capital. Here, we
  test whether researchers preferentially collaborate with same-gendered colleagues,
  using more stringent methods and a larger dataset than in past work. Our results
  reaffirm that researchers co-publish with colleagues of the same gender, and show
  that this 'gender homophily' is slightly stronger today than it was 10 years ago.
  Contrary to our expectations, we found no evidence that homophily is driven mostly
  by senior academics, and no evidence that homophily is strongest in fields where
  women are in the minority. Interestingly, journals with a high impact factor for their discipline tended to have comparatively low homophily, as predicted if mixed-gender teams produce better research.
  We discuss potential causes of gender homophily in science. \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining,
  Women in STEM. 
---

@@

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha' > 0$, and for many of these the FDR-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender.

@@

S1 Table. The $\alpha'$ values calculated for each journal, in the 2005 and 2015 samples, for each type of author (i.e. all authors, first authors, and last authors). The tables gives the impact factor of each journal, the sample size, $\alpha$ and $\alpha'$ and their 95% CIs, and the p-value from a 2-tailed test evaluating the null hypothesis that $\alpha$ is zero (both raw and FDR-corrected p-values are shown).

@@

>>>>

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers
csl: plos.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious
  biases, and often feel excluded from formal and informal opportunities for research
  collaboration. In addition to ensuring fairness and helping to redress gender imbalance
  in the academic workforce, increasing women's access to collaboration could help
  scientific progress by drawing on more of the available human capital. Here, we
  test whether researchers preferentially collaborate with same-gendered colleagues,
  using more stringent methods and a larger dataset than in past work. Our results
  reaffirm that researchers co-publish with colleagues of the same gender, and show
  that this 'gender homophily' is slightly stronger today than it was 10 years ago.
  Contrary to our expectations, we found no evidence that homophily is driven mostly
  by senior academics, and no evidence that homophily is strongest in fields where
  women are in the minority. Interestingly, journals with a high impact factor for their discipline tended to have comparatively low homophily, as predicted if mixed-gender teams produce better research.
  We discuss potential causes of gender homophily in science. \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining,
  Women in STEM. 
---

@@

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha' > 0$, and for many of these the FDR-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6; S1 Data). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender.

@@

S1 Table. Sample sizes for the two datasets, which comprise papers published in the timeframes August 2005 - August 2006, and August 2015 - August 2016.

@@

S1 Data: This spreadsheet shows the $\alpha$ values calculated for each journal, in the 2005 and 2015 samples, and for each type of author (all authors, first authors, and last authors). The tables gives the impact factor of each journal, the sample size, $\alpha$ and $\alpha'$ and their 95% CIs, and the p-value from a 2-tailed test evaluating the null hypothesis that $\alpha$ is zero (both raw and FDR-corrected p-values are shown).
